---
title: "'Generation Debt' Turns 40: Modeling Millennial Debt Trajectories in a Multilevel Framework"
author: "Ethan Weiland"
date: 2024-04-23
format: beamer
editor: visual
---

```{r, include=FALSE}
rm(list=ls()) # clearing environment

# Packages
library(tidyverse)
library(survey)
library(stargazer)
library(lme4)
library(lmerTest)

# Reading in data
load("total_debt.RData")

# Setting survey design
total_debt_design <- svydesign(data=total_debt, 
                               strata=~VSTRAT,
                               id=~VPSU,
                               nest=TRUE,
                               weights=~weight)
```

## Introduction

-   Widespread public concern over finances of young American adults ("Millennials")

    -   Especially regarding debt - labeled "Generation Debt" (Kamenetz 2006)

-   Research in sociology has confirmed high debt levels and financial distress for Millennials (Houle 2014)

-   Other researchers have explored implications of high debt levels, for example associations between debt and fertility (Nau, Dwyer, & Hodson 2015), physical health (Lippert, Houle, & Walsemann 2022), home buying (Houle and Berger 2015), and cohabitation (Addo 2014)

-   *Much of the sociological research studied Millennials in their twenties, what has changed now that Millennials are in their thirties?*

## Research Questions

1.  How much of the variance in debt is attributable to between-individual and within-individual differences?

2.  What is the association between debt and time?

3.  What is the association between parental net worth and initial debt levels and debt trajectories?

## Data

```{r, include=FALSE}
N <- length(unique(total_debt$id))
n <- nrow(total_debt)
```

-   National Longitudinal Survey of Youth 1997 (NLSY97)

    -   Nationally representative panel study of Americans between the ages of 12 and 16 in 1997

        -   Respondents are between ages of 39 and 43 today

    -   Currently in its 20th round

    -   Administered by the Bureau of Labor Statistics

-   Respondents are asked about debt every five years in Youth Asset (YAST) modules

    -   Currently six YAST modules - YAST20 through YAST40

-   Sample restricted to respondents reporting debt at least once

    -   Number of individuals (N) = `r N`

    -   Number of measurements (n) = `r n`

## Variables

-   Dependent Variable = Debt

    -   Includes automobile debt, student debt, debt owed to family/friends, credit card debt, store debt, healthcare debt, and other forms of debt

    -   Does [not]{.underline} include housing debt

-   Independent Variables

    -   Time = measurement occasion

    -   Parental Net Worth = household assets minus household debt of respondents' parents measured in Round 1 of survey (when respondent was between 12-16)

-   Control Variables

    -   Level-1 = age, education, income, assets, marital status, number of children

    -   Level-2 = sex, race, parent income, parent education

-   All financial measures adjusted for inflation and missing values imputed using bagging decision tree (machine learning algorithm)

## Methods

-   Implement a multilevel model with individuals as the level-2 units and measurement occasions as the level-1 units

-   RQ #1: How much of the variance in debt is attributable to between-individual and within-individual differences?

    -   Run a null model and calculate residual intraclass correlation coefficient

-   RQ #2: What is the association between debt and time?

    -   Test the fixed and random effect of time

-   RQ #3: What is the association between parental net worth and initial debt levels and debt trajectories?

    -   Test the fixed effect of parental net worth (impact on initial debt) and interaction between time and parental net worth (impact on debt trajectories)

## Total Debt Inferences

```{r, echo=FALSE, results="asis"}
means <- svyby(formula=~debt, # mean inference across YAST modules
               by=~time,
               design=total_debt_design,
               FUN=svymean,
               keep.names=FALSE)
variances <- svyby(formula=~debt, # variance inference across YAST modules
                   by=~time,
                   design=total_debt_design,
                   FUN=svyvar,
                   keep.names=FALSE)
variances <- variances[,-1]
variances <- sqrt(variances) # converting to standard deviations
table <- cbind(means, variances)
colnames(table) <- c("Age",
                      "Mean Debt",
                      "Mean Debt SE",
                      "Debt SD",
                      "Debt SD SE")
table$Age[table$Age==0] <- 20
table$Age[table$Age==1] <- 25
table$Age[table$Age==2] <- 30
table$Age[table$Age==3] <- 35
table$Age[table$Age==4] <- 40
table <- round(table, 2)

stargazer(table,
          type="latex",
          summary=FALSE,
          header=FALSE,
          rownames=FALSE,
          title=paste0("Total Debt Inferences Across Age, N=(",
                       N,
                       ")"),
          font.size="small",
          notes=c("Source: NLSY97. Among respondents reporting debt at least",
                  "once. Adjusted for survey design and inflation."))
```

## Null Model

```{r, echo=FALSE, results="asis"}
model1 <- lme4::lmer(debt_logged ~ 1 + # null model to calculate ICC
                       (1 | id),
                     data = total_debt)

# building stargazer() table to present results
mod_stats <- tribble(~stat, ~ model1,
                     "\\it{Level-Two Variation}", NA,
                     "$\\tau_{0}$ Intercept SD", round(as.numeric(attributes(VarCorr(model1)$"id")$stddev), 2),
                     "\\it{Level-One Variation}", NA,
                     "$\\sigma$ Residual SD", round(summary(model1)$sigma, 2),
                     "Number of Groups (Individuals)", N,
                     "Number of Observations (Measurements)", n,
                     "Deviance", round(deviance(model1, REML=F), 2))
stargazer(model1,
          type="latex",
          header=FALSE,
          font.size="small",
          title="Null Model for Total Debt",
          dep.var.labels=c("Total Debt (Logged)"),
          p=NA, # don't want p-values
          digits=2,
          omit.table.layout="s",
          add.lines=lapply(1:nrow(mod_stats), function(i) unlist(mod_stats[i, ])),
          notes=c("Source: NLSY97"),
          notes.append=F)
```

## Research Question #1

```{r, include=FALSE}
icc <- as.numeric(attributes(VarCorr(model1)$"id")$stddev)^2 / (summary(model1)$sigma^2 + as.numeric(attributes(VarCorr(model1)$"id")$stddev)^2)
icc <- round(icc, 2)
```

-   RQ #1: How much of the variance in debt is attributable to between-individual and within-individual differences?

-   The intraclass correlation coefficient (ICC) for the null model is `r icc`

-   `r icc*100`% of the variation in total debt is attributable to *between*-individual differences

## Model Formulation

Level 1:

$Debt_{ij} = \beta_{0j} + \beta_{1j}Time_{ij} + \beta_{2}Time^2_{ij} + \beta_{3}Age_{ij} + \beta_{4}Education_{ij} + \beta_{5}Income_{ij} + \beta_{6}Assets_{ij} + \beta_{7}Married_{ij} + \beta_{8}NumChildren_{ij} + \beta_{9}HasDebt_{ij} + r_{ij}$

Level 2:

$\beta_{0j} = \gamma_{00} + \gamma_{01}ParentNetWorth_j + \gamma_{02}ParentEducation_j + \gamma_{03}ParentIncome_j + \gamma_{04}Sex_j + \gamma_{05-07}Race_j + u_{0j}$

$\beta_{1j} = \gamma_{10} + \gamma_{11}ParentNetWorth_j$

Where

$r_{ij}$ \~ N(0,$\sigma^2$) and $\begin{pmatrix} u_{0j}\\u_{1j}\end{pmatrix}$ \~ *MVN* ($\begin{pmatrix} 0\\ 0 \end{pmatrix}$, $\begin{pmatrix}\tau^2_0 & \tau_{01} \\ \tau_{10} & \tau^2_1 \end{pmatrix}$)

```{r, include=FALSE, results="asis"}
# Function to automate calculating descriptives for continuous level-1 variables (loops over measurement occassions)
level_1_descriptives <- function(variable) {
  mean <- NA
  for (i in 0:max(total_debt$time)) {
    mean[i+1] <- mean(filter(total_debt, time==i)[[variable]])
  }
  sd <- NA
  for (i in 0:max(total_debt$time)) {
    sd[i+1] <- sd(filter(total_debt, time==i)[[variable]])
  }
  min <- NA
  for (i in 0:max(total_debt$time)) {
    min[i+1] <- min(filter(total_debt, time==i)[[variable]])
  }
  max <- NA
  for (i in 0:max(total_debt$time)) {
    max[i+1] <- max(filter(total_debt, time==i)[[variable]])
  }
  output <- cbind(mean, sd, min, max)
  return(output)
}

# Constructing table
table <- level_1_descriptives("debt") # dependent variable
table <- rbind(table, c(mean(total_debt$parent_net_worth), # parent net worth row
                        sd(total_debt$parent_net_worth),
                        min(total_debt$parent_net_worth),
                        max(total_debt$parent_net_worth)))
table <- rbind(table, c(mean(total_debt$parent_income), # parent income row
                        sd(total_debt$parent_income),
                        min(total_debt$parent_income),
                        max(total_debt$parent_income)))
table <- rbind(table, c(mean(total_debt$parent_education), # parent education row
                        sd(total_debt$parent_education),
                        min(total_debt$parent_education),
                        max(total_debt$parent_education)))
table <- rbind(table, c(sum(total_debt$sex=="Female")/nrow(total_debt), # sex row
                        NA,
                        0,
                        1))

# Race factor
table <- rbind(table, c(sum(total_debt$race=="Non-Black / Non-Hispanic")/nrow(total_debt),
                        NA,
                        0,
                        1))
table <- rbind(table, c(sum(total_debt$race=="Black")/nrow(total_debt),
                        NA,
                        0,
                        1))
table <- rbind(table, c(sum(total_debt$race=="Hispanic")/nrow(total_debt),
                        NA,
                        0,
                        1))
table <- rbind(table, c(sum(total_debt$race=="Multiracial (Non-Hispanic)")/nrow(total_debt),
                        NA,
                        0,
                        1))

table <- rbind(table, level_1_descriptives("age")) # age rows
table <- rbind(table, level_1_descriptives("educ")) # education rows
table <- rbind(table, level_1_descriptives("income")) # income rows
table <- rbind(table, level_1_descriptives("assets")) # assets rows

# Marital status indicator
table <- rbind(table, c(sum(filter(total_debt, time==0)$marital_status=="Married")/nrow(filter(total_debt, time==0)),
                        NA,
                        0,
                        1)) 
table <- rbind(table, c(sum(filter(total_debt, time==1)$marital_status=="Married")/nrow(filter(total_debt, time==1)),
                        NA,
                        0,
                        1))
table <- rbind(table, c(sum(filter(total_debt, time==2)$marital_status=="Married")/nrow(filter(total_debt, time==2)),
                        NA,
                        0,
                        1))
table <- rbind(table, c(sum(filter(total_debt, time==3)$marital_status=="Married")/nrow(filter(total_debt, time==3)),
                        NA,
                        0,
                        1))
table <- rbind(table, c(sum(filter(total_debt, time==4)$marital_status=="Married")/nrow(filter(total_debt, time==4)),
                        NA,
                        0,
                        1))

table <- rbind(table, level_1_descriptives("num_children")) # number of children row

# Has debt indicator
table <- rbind(table, c(sum(filter(total_debt, time==0)$has_debt=="1")/nrow(filter(total_debt, time==0)),
                        NA,
                        0,
                        1)) 
table <- rbind(table, c(sum(filter(total_debt, time==1)$has_debt=="1")/nrow(filter(total_debt, time==1)),
                        NA,
                        0,
                        1)) 
table <- rbind(table, c(sum(filter(total_debt, time==2)$has_debt=="1")/nrow(filter(total_debt, time==2)),
                        NA,
                        0,
                        1)) 
table <- rbind(table, c(sum(filter(total_debt, time==3)$has_debt=="1")/nrow(filter(total_debt, time==3)),
                        NA,
                        0,
                        1)) 
table <- rbind(table, c(sum(filter(total_debt, time==4)$has_debt=="1")/nrow(filter(total_debt, time==4)),
                        NA,
                        0,
                        1)) 

table <- round(table, 2)

# Labeling
varnames <- c("Debt 20 ($)",
              "Debt 25 ($)",
              "Debt 30 ($)",
              "Debt 35 ($)",
              "Debt 40 ($)",
              "Parent Net Worth ($1000s)",
              "Parent Income ($1000s)",
              "Parent Education",
              "Sex (1 = Female)",
              "Non-Black / Non-Hispanic",
              "Black",
              "Hispanic",
              "Multiracial (Non-Hispanic)",
              "Age 20",
              "Age 25",
              "Age 30",
              "Age 35",
              "Age 40",
              "Education 20",
              "Education 25",
              "Education 30",
              "Education 35",
              "Education 40",
              "Income ($1000s) 20",
              "Income ($1000s) 25",
              "Income ($1000s) 30",
              "Income ($1000s) 35",
              "Income ($1000s) 40",
              "Assets ($1000s) 20",
              "Assets ($1000s) 25",
              "Assets ($1000s) 30",
              "Assets ($1000s) 35",
              "Assets ($1000s) 40",
              "Married 20",
              "Married 25",
              "Married 30",
              "Married 35",
              "Married 40",
              "# Children 20",
              "# Children 25",
              "# Children 30",
              "# Children 35",
              "# Children 40",
              "Has Debt 20",
              "Has Debt 25",
              "Has Debt 30",
              "Has Debt 35",
              "Has Debt 40")
table <- cbind(varnames, table)
colnames(table) <- c("Variable", "Mean", "StdDev", "Min.", "Max.")
stargazer(table,
          type="latex",
          summary=FALSE,
          header=FALSE,
          rownames=FALSE,
          title=paste0("Descriptive Statistics, N=(",
                       N,
                       ")"),
          font.size="small",
          notes=c("Source: NLSY97. Among respondents reporting debt at least",
                  "once. Adjusted for inflation."))

# Ran into difficulty rendering onto slide, so rendered as PDF page and manually inserted into slide deck
  # Important LATEX commands to format in PDF nicely:
  # \setcounter{table}{2}
  # \vspace{-100pt}
  # \pagenumbering{gobble}

```

```{r, include=FALSE}
model2 <- lme4::lmer(debt_logged ~ 1 +
                       time +
                       time_sq +
                       parent_net_worth +
                       time:parent_net_worth +
                       parent_education +
                       parent_income +
                       sex +
                       race +
                       age +
                       educ +
                       income +
                       assets +
                       marital_status +
                       num_children +
                       has_debt +
                       (1 + time| id),
                     data = total_debt,
                     control = lmerControl(optimizer ="Nelder_Mead")) # model won't converge
                                                                      # with default optimizer

# building stargazer() table to present results
mod_stats <- tribble(~stat, ~ model2,
                     "\\it{Level-Two Variation}", NA,
                     "$\\tau_{0}$ Intercept SD", round(as.numeric(attributes(VarCorr(model2)$"id")$"stddev"[1]), 2),
                     "$\\tau_{1}$ Slope SD", round(as.numeric(attributes(VarCorr(model2)$"id")$"stddev"[2]), 2),
                     "$\\tau_{01}$ Intercept Slope Correlation", round(as.numeric(attributes(VarCorr(model2)$"id")$"correlation"[2,1]), 2),
                     "\\it{Level-One Variation}", NA,
                     "$\\sigma$ Residual SD", round(summary(model2)$sigma, 2),
                     "Number of Groups (Individuals)", N,
                     "Number of Observations (Measurements)", n,
                     "Deviance", round(deviance(model2, REML=F), 2))

stargazer(model2,
          type="latex",
          header=FALSE,
          font.size="scriptsize",
          title="Longitudinal Random Slope Model for Total Debt (Logged)",
          dep.var.labels=c("Total Debt (Logged)"),
          covariate.labels=c("\\gamma_{10} Time",
                             "\\beta_{2} Time^2",
                             "\\gamma_{01} Parent Net Worth ($1000s)",
                             "\\gamma_{02} Parent Education",
                             "\\gamma_{03} Parent Income (\\$1000s)",
                             "\\gamma_{04} Sex (Female=1)",
                             "\\gamma_{05} Black",
                             "\\gamma_{06} Hispanic",
                             "\\gamma_{07} Multiracial (Non-Hispanic)",
                             "\\beta_{3} Age",
                             "\\beta_{4} Education",
                             "\\beta_{5} Income (\\$1000s)",
                             "\\beta_{6} Assets (\\$1000s)",
                             "\\beta_{7} Married",
                             "\\beta_{8} # Children",
                             "\\beta_{9} Has Debt",
                             "\\gamma_{11} Time:Parent Net Worth"),
          digits=2,
          p=rep(NA, nrow(summary(model2)$coefficients)), # don't want p-values
          omit.table.layout="s",
          add.lines=lapply(1:nrow(mod_stats), function(i) unlist(mod_stats[i, ])),
          notes=c("Source: NLSY97"),
          notes.append=FALSE)

# Ran into difficulty rendering onto slide, so rendered as PDF page and manually inserted into slide deck
  # Important LATEX commands to format in PDF nicely:
  # \setcounter{table}{3}
  # \vspace{-60pt}
  # \pagenumbering{gobble}
```

## Model Assumptions

```{r, echo=FALSE}
ggplot(data.frame(reffect=ranef(model2)$id[,1]), aes(x=reffect)) +
  geom_histogram(bins=30) +
  labs(title=expression("Figure 1. Distribution of"~U[0]),
       y="Count",
       x=expression(U[0]))
```

## Model Assumptions

```{r, echo=FALSE}
ggplot(data.frame(reffect=ranef(model2)$id[,2]), aes(x=reffect)) +
  geom_histogram(bins=30) +
  labs(title=expression("Figure 2. Distribution of"~U[1]),
       y="Count",
       x=expression(U[1]))
```

## Research Question #2

```{r, include=FALSE}
model2 <- lmerTest::lmer(debt_logged ~ 1 + # to get p-values
                           time +
                           time_sq +
                           parent_net_worth +
                           time:parent_net_worth +
                           parent_education +
                           parent_income +
                           sex +
                           race +
                           age +
                           educ +
                           income +
                           assets +
                           marital_status +
                           num_children +
                           has_debt +
                           (1 + time | id),
                         data = total_debt,
                         REML=TRUE,
                         control = lmerControl(optimizer = "Nelder_Mead"))

coefficients_table <- summary(model2)$coefficients[,]
colnames(coefficients_table) <- attr(summary(model2)$coefficients, "dimnames")[[2]]
rownames(coefficients_table) <- attr(summary(model2)$coefficients, "dimnames")[[1]]

gamma10 <- coefficients_table[which(rownames(coefficients_table)=="time"),
                              which(colnames(coefficients_table)=="Estimate")]
beta2 <- coefficients_table[which(rownames(coefficients_table)=="time_sq"),
                             which(colnames(coefficients_table)=="Estimate")]
```

-   RQ #2: What is the association between debt and time?

-   $\gamma_{10}$ = `r round(gamma10, 5)` (p \< 0.001)

    -   Time is significantly associated with total debt

    -   The average growth rate of total debt across each measurement occasion (5 years) is `r round(gamma10, 2)*100`%

-   $\beta_2$ = `r round(beta2, 5)` (p \< 0.001)

    -   Time$^2$ is significantly associated with total debt

    -   On average, the growth rate of total debt is *decelerating*

## Research Question #2

```{r, include=FALSE}
model2_null <- lmerTest::lmer(debt_logged ~ 1 + # to get p-values
                                         time +
                                         time_sq +
                                         parent_net_worth +
                                         time:parent_net_worth +
                                         parent_education +
                                         parent_income +
                                         sex +
                                         race +
                                         age +
                                         educ +
                                         income +
                                         assets +
                                         marital_status +
                                         num_children +
                                         has_debt +
                                         (1 | id), # no random slope
                                       data = total_debt,
                                       REML = TRUE,
                                       control = lmerControl(optimizer = "Nelder_Mead"))
-2*(logLik(model2_null, REML=T)-logLik(model2, REML=T)) > qchisq(.05, 2, lower.tail=FALSE) 
  # ^ reject the reduced model = support for random slope
AIC(model2) < AIC(model2_null) # support for random slope
BIC(model2) < BIC(model2_null) # support for random slope

tau1 <- round(as.numeric(attributes(VarCorr(model2)$"id")$"stddev"[2]), 2)
tau01 <- as.numeric(attributes(VarCorr(model2)$"id")$"correlation"[2,1])
```

-   Likelihood Ratio Test, AIC, and BIC all prefer the model with random slope for time

    -   There is significant variation in the growth rate of debt

-   $\tau_1$ = `r round(tau1, 2)`

    -   For 95% of individuals, the growth rate of debt varies by $\tau_1 * 4$ = `r round((tau1*4)*100)`%

-   $\tau_{01}$ = `r round(tau01, 5)`

    -   There is [no]{.underline} correlation between initial debt levels and the rate of growth

## Research Question #3

```{r, include=FALSE}
gamma01 <- coefficients_table[which(rownames(coefficients_table)=="parent_net_worth"),
                              which(colnames(coefficients_table)=="Estimate")]
gamma11 <- coefficients_table[which(rownames(coefficients_table)=="time:parent_net_worth"),
                              which(colnames(coefficients_table)=="Estimate")]
```

-   RQ #3: What is the association between parental net worth and initial debt levels and debt trajectories?

-   $\gamma_{01}$ = `r round(gamma01, 5)` (p \< 0.01)

    -   An increase of \$1,000 in parental net worth decreases initial debt levels by `r round(gamma01*-100, 3)`%

        OR

    -   An increase of \$100,000 in parental net worth decreases initial debt levels by `r round(gamma01*-10000, 2)`%

-   $\gamma_{11}$ = `r round(gamma11, 6)` (p \< 0.001)

    -   An increase in parental net worth decreases the growth rate of total debt

-   While statistically significant, these associations are not *practically* significant

## Limitations

-   A mixture model (e.g., zero-inflated model) would better handle respondents with \$0 debt

-   Multilevel packages in R like `lme4` don't include options for specifying survey design and the `survey` package doesn't include options for multilevel modeling

-   NLSY97 poorly measures sex/gender and race

## Conclusion

-   Among Millennials...

    -   Most of the variation in debt comes from within individuals rather than between individuals

    -   Debt significantly increases over time, although the increase diminishes

    -   Individuals experience varied trajectories of debt growth

    -   Parental net worth does not meaningfully impact debt

## References

1.  Addo, Fenaba R. 2014. "Debt, Cohabitation, and Marriage in Young Adulthood." Demography 51(5):1677--1701. doi: 10.1007/s13524-014-0333-6.

2.  Houle, Jason N. 2014. "A Generation Indebted: Young Adult Debt across Three Cohorts." Social Problems 61(3):448--65. doi: 10.1525/sp.2014.12110.

3.  Kamenetz, Anya. 2006. Generation Debt: How Our Future Was Sold Out for Student Loans, Bad Jobs, No Benefits, and Tax Cuts for Rich Geezers--And How to Fight Back. New York, New York: Riverhead Books.

4.  Lippert, Adam M., Jason N. Houle, and Katrina M. Walsemann. 2022. "Student Debt and Cardiovascular Disease Risk Among U.S. Adults in Early Mid-Life." American Journal of Preventive Medicine 63(2):151--59. doi: 10.1016/j.amepre.2022.02.002.

5.  Nau, Michael, Rachel E. Dwyer, and Randy Hodson. 2015. "Can't Afford a Baby? Debt and Young Americans." Research in Social Stratification and Mobility 42:114--22. doi: 10.1016/j.rssm.2015.05.003.

## Thank You!

<https://github.com/ethanphilipweiland/millennial-debt>
